---
title: "Create manifest"
output:
  html_document:
    df_print: paged
---
```{r}
library("tidyverse")
library("knitr")
library("here")
```

To import demultiplexed paired-end reads into Qiime2, first we need to create a manifest to link toghter the metadata with the Illumina samples. More infor can be found here [Qiime2 import tutorial] (https://docs.qiime2.org/2018.8/tutorials/importing/), see “Fastq manifest”.

The final format should look like this.

```{r eval=FALSE, include=TRUE}

sample-id,absolute-filepath,direction
sample-1,$PWD/some/filepath/sample1_R1.fastq,forward
sample-1,$PWD/some/filepath/sample1_R2.fastq,reverse
```

We have four type of fastq-names:

801316623959_S44_L001_R1_001.fastq.gz
verdi-1-171109_S110_L001_R1_001.fastq.gz
Mock-2_S114_L001_R1_001.fastq.gz
Undetermined_S0_L001_R1_001.fastq.gz

**The following code is not general, this is specific to the VERDI project sample naming**

```{r echo=TRUE}
# Createa a dataframe based on 
create_data_frame <- function(regex_pattern){
  list_of_files <- 
      list.files(path = "../../../../data/raw/", pattern = regex_pattern )
  df <- 
    data.frame(
      sample_id = list_of_files %>% as.character(),
      absolute_filepath = paste("$PWD/../../data/raw/", list_of_files, sep = ""),
      direction = sapply(list_of_files, direction),
      stringsAsFactors = FALSE,
      row.names = 1:(list_of_files %>% length())) 
}

# Get the direction dependet on _R1_ or __R2__
direction <- function(x){
  ifelse(grepl(pattern = "_R1_", x), "forward", "reverse")
}

names_df <- create_data_frame("(^[0-9]{4})")
names_df$sample_id <- sapply(names_df$sample_id, substr, 5, 12)

verdi_df <- create_data_frame("(verdi)")
verdi_df$sample_id <- sapply(strsplit(verdi_df$sample_id, "_"), "[", 1)

mock_df <- create_data_frame("(Mock)")
mock_df$sample_id <- sapply(strsplit(mock_df$sample_id, "-"), "[", 1)

undetermined_df <- create_data_frame("(Undetermined)")
undetermined_df$sample_id <- sapply(strsplit(undetermined_df$sample_id, "_"), "[", 1)

samples_df <- rbind(names_df, verdi_df, mock_df, undetermined_df)
kable(samples_df[c(1:3, 225:230),], align = "l", format = "html")

write.table(samples_df,
          file = "../../../verdi_snakemake/manifest_file",
          quote = FALSE,
          sep = ",",
          row.names = FALSE,
          col.names = FALSE)
```

